Best Practices for Alchemical Free Energy Calculations: v\versionnumber

### 2020-01-16 paper/manuscript.tex
Working my way through section 7, section 8 is done.

In the remainder of this paper, we will cover topics that are key to the preparation~(\ref{sec:step1}), choice and use of correct protocols~(\ref{sec:step2}), and finally the best practices that should be used for the analysis of alchemical calculations~(\ref{sec:data_analysis}). 

@@

The alchemical free energy protocols, that are discussed in Section~\ref{sec:step2}, are defined for a specific type of free energy calculation, i.e. a free energy of binding or a free energy of hydration. Depending on the type of simulations different considerations have to be made for ligands, solvent, and host molecules (in the case of the estimation of free energies of binding).

@@

For binding free energy calculations using a congeneric set of ligands and therefore likely to be most suited to a relative free energy protocol (see~\ref{sec:step2}). Other than the choice of forcefield for the parameterization of the ligands, some care has to be taken selecting binding poses for these ligands. Generally a common assumption for a congeneric series is that the binding mode is conserved. Therefore, if an X-ray structure of one of the ligands is available, this should be used to position the ligands in the putative binding site in an energetically reasonable conformation that entails no steric or electrostatic mismatch with the receptor. Checking the X-ray structure versus the experimental electron densities is important, as the resolution of part of the ligand or important sidechains may be based on the interpretation of the crystallographer rather than the available electron density. For example, looking at a cyclohexane ring density, a chair conformation is vastly more likely than that of a boat and may just be poorly assigned. 

@@

\section{What simulation protocol should I choose? -- Step 2}
\label{sec:step2}
Alchemical free energy calculations can be grouped into two main categories, ``absolute'' (\ref{fig:fig_absolute_thermodynamic_cycle}) and ``relative'' \footnote{The distinction is a bit of a misnomer, since both compute ratios of partition functions relative to another state, and neither computes an absolute free energy.}(\ref{fig:fig_binding_thermodynamic_cycle}), which differ in whether they compute properties for a single molecule (absolute) or compare properties of different, usually closely related, molecules (relative).

@@

Many of the issues of setup and protocol for alchemical calculations are common, but some are different between absolute and relative calculations, so before treating the common elements we consider the protocol differences.

\subsubsection{Choices unique to relative free energy calculations}
\label{sec:relative-fe-protocol}

\paragraph{Topologies} A critical first step in relative calculations is to select an approach to these calculations, determining whether to use a \emph{dual topology}, \emph{single topology}, or \emph{hybrid topology} approach to relative calculations.

@@

In single topology calculations, the overall transformation is set up to involve as few additional atoms as possible, so benzene would be typically changed into benzyl alcohol by changing one of the hydrogens into a carbon. To this new carbon you also find non-interacting atoms called ``dummy atoms'' (retaining their bonded interactions but not interacting with the rest of the system), which will become the two hydrogens attached to the new carbon and the OH group. Bond parameters as well as partial charges are adjusted accordingly. 
Thus, in a single topology calculation, atoms may change their type so relatively few dummy atoms are created. This is illustrated in left arm of Fig.~\ref{fig:fig_topology}. 
In contrast, in a dual topology free energy calculation, no atoms are allowed to change type~\cite{shirts2012best}. This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxy methyl group, then passing through an intermediate state where atoms which are becoming dummy atoms or ceasing being dummy atoms are partially interacting (this state may or may not be well defined~\cite{mobley2014blind}  and culminating in a state where benzyl alcohol is present along with the additional dummy atoms which was previously a corresponding hydrogen of the benzene. Fig.~\ref{fig:fig_topology}'s right branch depicts how such a dual topology works. 

@@

    \caption{\textbf{Two common topologies for alchemical calculations} left: A single topology uses softcore potentials to convert from one type of atom to an other. Dummy atoms (Du) are used when there is no corresponding maximum common substructure match between the two molecules for certain atoms. right: The dual topology does not convert one species to another, but only converts between Du atoms and an interacting species, but usually uses softcore potentials for this. The 'mixed' intermediate atoms are used in both dual and single topology approaches. Only the way the transformation occurs and the end states differ. Following the arrow along the left and right illustrate the differences.  Figure adapted from \url{http://www.alchemistry.org/wiki/Constructing_a_Pathway_of_Intermediate_States}}

@@

At present, the most widely used approaches, such as in Schr\"{o}dinger's FEP+~\cite{wang2019protein} and in FESetup~\cite{loeffler2015fesetup} (for which calculations may be planned with Lead Optimization Mapper (LOMAP)~\cite{liu2013lead} seem to use single topology approaches, though some codes only support dual topology.

@@

Rigorously, this process essentially comprises a maximal common substructure (MCSS) search of the molecules involved to identify the common substructure -- though the parameters of the MCSS search will differ depending on whether single or dual topology calculations are planned.

@@

    \caption{\textbf{Illustration of maximum common substructure matches} MCSS is shown in green for when A: a restrictive MCSS match is used and in B: ring breaking is allowed.}

@@

There are different tools that allow the generation of MCSS matches as well as single topology input. A large number of software can compute MCSS matches, some rely on RDKit, such as LOMAP or FESetup and partially BioSimSpace, others such as fkckombu are standalone tools~\cite{rdkit2019Dec, loeffler2015fesetup, liu2013lead, hedges2019biosimspace, kawabata20143d}. Schr\"{o}dinger's FEP+ planning tool is based on a version of LOMAP, and it also uses MCSS matching as well as 3D considerations to plan the network of single topology calculations between molecules~\cite{wang2015accurate}. 
MCSS searches can be relatively time consuming, so if the goal is to assess a library of ligands to identify promising pairs for relative calculations, it can be helpful to use faster approaches such as shape similarity to perform an initial assessment and then use MCSS only to identify final mappings for relative calculations.
The MCSS approach, though relatively standard, takes into account only topological similarity.
It is possible that changes in binding mode could actually require a different choice of mapping, so in some cases mappings may need to be planned differently depending on 3D positioning of atoms in space.
Single topology relative calculations, and calculations based on substructure searches, only work if in fact the ligands share a common substructure, e.g. are part of a congeneric series.

@@

\paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic, in part because relative calculations rely on the free energy contributions of dummy atoms canceling between different legs of the thermodynamic cycle~\cite{}, which may not be true whenever dummy atoms are involved in rings.

@@

Methods in experimental design have been applied to the construction of the perturbation maps.  Yang et al.~\cite{yang2020optimal} optimized the perturbation map by selecting a fixed number of calculations from the pairwise perturbations so that the resulting set of calculations minimize the total variance.  Xu~\cite{xu2019diffnet} optimized the perturbation map by allocating different amounts of simulation time to different pairwise perturbations so as to minimize the total variance, given the total simulation time of all the perturbation calculations.  Both approaches lead to substantial reduction in the statistical error of the estimated free energies.  

@@

% \begin{Checklists*}[p!]
% \begin{checklist}{Absolute simulations}
% \textbf{What are the main things I need to consider for an absolute alchemical free energy calculation?}
% Extensive explanation for the checklist questions can be found in Section~\ref{sec:step2}.
% \begin{itemize}
% \item Topology
% \item Restraints
% \item Standard state handling
% \end{itemize}
% \end{checklist}

% \begin{checklist}{Relative simulations}
% \textbf{What are the main things I need to consider for an relative alchemical free energy calculation?}
% Extensive explanation for the checklist questions can be found in Section~\ref{sec:step2}.
% \begin{itemize}
% \item First thing
% \item Also remember
% \item And finally
% \end{itemize}
% \end{checklist}

@@

\section*{Author Contributions}
%%%%%%%%%%%%%%%%
% This section mustt describe the actual contributions of
% author. Since this is an electronic-only journal, there is
% no length limit when you describe the authors' contributions,
% so we recommend describing what they actually did rather than
% simply categorizing them in a small number of
% predefined roles as might be done in other journals.
%
% See the policies ``Policies on Authorship'' section of https://livecoms.github.io
% for more information on deciding on authorship and author order.
%%%%%%%%%%%%%%%%
\todo[inline, color={green!20}]{ASJSM: @everyone, can you check your contribution is acknowledge and if not add the appropriate information. }
(Explain the contributions of the different authors here)
ASJSM: Coordinated the document, contributed to most sections, and co-designed figures ~\ref{} and created figures \ref{fig:fig_what_is_lambda},~\ref{fig:fig1_what_is_alchemy},~\ref{fig:overlap} and replotted~\ref{fig:automatic-equilibration-detection} and improved ~\ref{fig:fig_types_of_networks}.
JS: created figures \ref{fig:fig_what_is_alchemy},~\ref{fig:fig_binding_thermodynamic_cycle},~\ref{fig:fig_topology},~\ref{fig:fig_mcss},~\ref{fig:fig_types_of_networks},~\ref{fig:fig_absolute_thermodynamic_cycle},~\ref{fig:scatterplot_analysis}, the data representation section, the checklist and contributed to general formatting discussions and editing.
MK: Contributed to the Data Analysis section, provided the data for figure~\ref{fig:overlap} and helped edit the paper.
DLM: Contributed to the outline, drafted some of the sections, gave ideas on figures, and helped edit the paper.
GT: Contributed to introduction and drug discovery sections, and helped edit the paper.
AR: Created figure~\ref{fig:freeenergytrajectories}, contributed to section \ref{sec:step2}, and helped edit the paper.
MRS: Helped create figure~\ref{fig:fig_what_is_lambda}, wrote sections describing choices for alchemical pathways and analysis for free energy calculations.
LNN: Helped write the simulation length, stopping conditions, and information saving section.
BA: Helped write the uncertainty estimation, stopping conditions, and output analysis sections and created figure ~\ref{fig:convergence_forward_reverse}.
JM: Contributed to sections~\ref{subsec:reproducible},~\ref{sec:step1},~\ref{} 4.2, 5.3, 6, 8.3.1, 8.3.2\
JDC: Wrote section~\ref{sec:decorrelating-samples} and~\ref{sec:automatic-equilibration-detection} discussed structure and design of the whole document, suggested Figs.~\ref{fig:fig_what_is_alchemy} and~\ref{fig:fig_sampling_scheme} \todo[inline, color={red!20}]{@JDC check info}
HX: Contributed section~\ref{subsec:accuracy}, to section~\ref{sec:relative-fe-protocol}, and to section~\ref{sec:are-they-good}.

>>>>

In the remainder of this paper, we will cover topics that are key to the preparation~(\ref{sec:step1}), choice and use of correct protocols~(\ref{sec:simulation_protocol_choice}), and finally the best practices that should be used for the analysis of alchemical calculations~(\ref{sec:data_analysis}). 

@@

The alchemical free energy protocols, that are discussed in Section~\ref{sec:simulation_protocol_choice}, are defined for a specific type of free energy calculation, i.e. a free energy of binding or a free energy of hydration. Depending on the type of simulations different considerations have to be made for ligands, solvent, and host molecules (in the case of the estimation of free energies of binding).

@@

For binding free energy calculations using a congeneric set of ligands and therefore likely to be most suited to a relative free energy protocol (see~\ref{sec:simulation_protocol_choice}). Other than the choice of forcefield for the parameterization of the ligands, some care has to be taken selecting binding poses for these ligands. Generally a common assumption for a congeneric series is that the binding mode is conserved. Therefore, if an X-ray structure of one of the ligands is available, this should be used to position the ligands in the putative binding site in an energetically reasonable conformation that entails no steric or electrostatic mismatch with the receptor. Checking the X-ray structure versus the experimental electron densities is important, as the resolution of part of the ligand or important sidechains may be based on the interpretation of the crystallographer rather than the available electron density. For example, looking at a cyclohexane ring density, a chair conformation is vastly more likely than that of a boat and may just be poorly assigned. 

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% What simulation protocol should I choose   %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{What simulation protocol should I choose?}
\label{sec:simulation_protocol_choice}
Alchemical free energy calculations can be grouped into two main categories, ``absolute'' (see Fig.~\ref{fig:fig_absolute_thermodynamic_cycle}) and ``relative'' \footnote{The distinction is a bit of a misnomer, since both compute ratios of partition functions relative to another state, and neither computes an absolute free energy.}(see Fig.~\ref{fig:fig_binding_thermodynamic_cycle}), which differ in whether they compute properties for a single molecule (absolute) or compare properties of different, usually closely related, molecules (relative).

@@

Many of the issues around simualtion setup and protocol choice for alchemical calculations are common, but there are some differences between absolute and relative calculations. First we will consider protocol differences  so before treating the common elements.
\subsubsection{Choices unique to relative free energy calculations}
\label{sec:relative-fe-protocol}
\paragraph{Topologies} 
A critical first step in relative calculations is to select an approach to these calculations, determining whether to use a \emph{dual topology}, \emph{single topology}, or \emph{hybrid topology} approach to relative calculations.

@@

In single topology calculations, the overall transformation is set up to involve as few additional atoms as possible, so benzene would be typically changed into benzyl alcohol by changing one of the hydrogens into a carbon. On this new carbon you will also find non-interacting atoms called ``dummy atoms'' (retaining their bonded interactions but not interacting with the rest of the system), which will become the two hydrogens attached to the new carbon and the OH group after the transformation. Bond parameters as well as partial charges are adjusted accordingly. 
Thus, in a single topology calculation, atoms may change their type ensuring only relatively few dummy atoms are created. This is illustrated in left arm of Fig.~\ref{fig:fig_topology}. 
In contrast, in a dual topology alchemical free energy calculation, no atoms are allowed to change type~\cite{shirts2012best}. This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxy methyl group, then passing through an intermediate state where atoms which are becoming dummy atoms or ceasing being dummy atoms are partially interacting (this state may or may not be well defined~\cite{mobley2014blind}  and culminating in a state where benzyl alcohol is present along with the additional dummy atoms which was previously a corresponding hydrogen of the benzene. Fig.~\ref{fig:fig_topology}'s right branch depicts how such a dual topology works. 

@@

    \caption{\textbf{Two common topologies for alchemical calculations: single and dual topology} \textbf{left}: A single topology uses softcore potentials to convert from one type of atom to an other. Dummy atoms (Du) are used when there is no corresponding maximum common substructure match between the two molecules for certain atoms. \textbf{right}: The dual topology does not convert one species to another, but only converts between Du atoms and an interacting species, but usually uses softcore potentials for this. The 'mixed' intermediate atoms are used in both dual and single topology approaches. Only the way the transformation occurs and the end states differ. Following the arrow along the left and right illustrate the differences.  Figure adapted from \url{http://www.alchemistry.org/wiki/Constructing_a_Pathway_of_Intermediate_States}}

@@

At present, the most widely used approaches, such as in Schr\"{o}dinger's FEP+~\cite{wang2015accurate,wang2019protein} and in BioSimSpace~\cite{hedges2019biosimspace} (for which calculations may be planned with Lead Optimization Mapper (LOMAP)~\cite{liu2013lead} seem to use single topology approaches, though some codes only support dual topology.

@@

Rigorously, this process comprises of a maximal common substructure (MCSS) search of the molecules involved to identify the common substructure -- though the parameters of the MCSS search will differ depending on whether single or dual topology calculations are planned.

@@

    \caption{\textbf{Illustration of maximum common substructure matches} MCSS is shown in green for when (\textbf{A}) a restrictive MCSS match is used and in (\textbf{B}) ring breaking is allowed.}

@@

There are different tools that allow the generation of MCSS matches as well as single topology input. A large number of software can compute MCSS matches, some rely on RDKit~\cite{rdkit2019Dec}, such as LOMAP~\cite{liu2013lead} or FESetup~\cite{loeffler2015fesetup} and partially BioSimSpace~\cite{hedges2019biosimspace}, others such as fkckombu~\cite{kawabata20143d} are standalone tools. Schr\"{o}dinger's FEP+ planning tool is based on a version of LOMAP, and it also uses MCSS matching as well as 3D considerations to plan the network of single topology calculations between molecules~\cite{wang2015accurate}. 
MCSS searches can be relatively time consuming, so if the goal is to assess a library of ligands to identify promising pairs for relative calculations, it can be helpful to use faster approaches such as shape similarity to perform an initial assessment and then use MCSS only to identify final mappings for relative calculations~\cite{raymond2002maximum,klabunde2012mars,jones2009elucidating}.
The MCSS approach, though relatively standard, takes into account only topological similarity. It is possible that changes in binding mode could actually require a different choice of mapping, so in some cases mappings may need to be planned differently depending on 3D positioning of atoms in space.
Single topology relative calculations, and calculations based on substructure searches, only work if in fact the ligands share a common substructure, e.g. are part of a congeneric series, see Fig.~\ref{fig:fig_mcss}

@@

\paragraph{Ring breaking and forming.} Relative free energy calculations for ring breaking and forming are particularly challenging/problematic (see Fig.~\ref{fig:fig_mcss} \textbf{B}), in part because relative calculations rely on the free energy contributions of dummy atoms canceling between different legs of the thermodynamic cycle, which may not be true whenever dummy atoms are involved in rings.

@@

Methods in experimental design have been applied to the construction of the perturbation maps.  Yang et al.~\cite{yang2020optimal} optimized the perturbation map by selecting a fixed number of calculations from the pairwise perturbations so that the resulting set of calculations minimize the total variance.  Xu~\cite{xu2019optimal} optimized the perturbation map by allocating different amounts of simulation time to different pairwise perturbations so as to minimize the total variance, given the total simulation time of all the perturbation calculations.  Both approaches lead to substantial reduction in the statistical error of the estimated free energies.  

@@

% \begin{Checklists*}[p!]
% \begin{checklist}{Absolute simulations}
% \textbf{What are the main things I need to consider for an absolute alchemical free energy calculation?}
% Extensive explanation for the checklist questions can be found in Section~\ref{sec:simulation_protocol_choice}.
% \begin{itemize}
% \item Topology
% \item Restraints
% \item Standard state handling
% \end{itemize}
% \end{checklist}

% \begin{checklist}{Relative simulations}
% \textbf{What are the main things I need to consider for an relative alchemical free energy calculation?}
% Extensive explanation for the checklist questions can be found in Section~\ref{sec:simulation_protocol_choice}.
% \begin{itemize}
% \item First thing
% \item Also remember
% \item And finally
% \end{itemize}
% \end{checklist}

@@

\section*{Author Contributions}
%%%%%%%%%%%%%%%%
% This section mustt describe the actual contributions of
% author. Since this is an electronic-only journal, there is
% no length limit when you describe the authors' contributions,
% so we recommend describing what they actually did rather than
% simply categorizing them in a small number of
% predefined roles as might be done in other journals.
%
% See the policies ``Policies on Authorship'' section of https://livecoms.github.io
% for more information on deciding on authorship and author order.
%%%%%%%%%%%%%%%%
\todo[inline, color={green!20}]{ASJSM: @everyone, can you check your contribution is acknowledge and if not add the appropriate information. }
(Explain the contributions of the different authors here)
ASJSM: Coordinated the document, contributed to most sections, and co-designed figures ~\ref{} and created figures \ref{fig:fig_what_is_lambda},~\ref{fig:fig1_what_is_alchemy},~\ref{fig:overlap} and replotted~\ref{fig:automatic-equilibration-detection} and improved ~\ref{fig:fig_types_of_networks}.
JS: created figures \ref{fig:fig_what_is_alchemy},~\ref{fig:fig_binding_thermodynamic_cycle},~\ref{fig:fig_topology},~\ref{fig:fig_mcss},~\ref{fig:fig_types_of_networks},~\ref{fig:fig_absolute_thermodynamic_cycle},~\ref{fig:scatterplot_analysis}, the data representation section, the checklist and contributed to general formatting discussions and editing.
MK: Contributed to the Data Analysis section, provided the data for figure~\ref{fig:overlap} and helped edit the paper.
DLM: Contributed to the outline, drafted some of the sections, gave ideas on figures, and helped edit the paper.
GT: Contributed to introduction and drug discovery sections, and helped edit the paper.
AR: Created figure~\ref{fig:freeenergytrajectories}, contributed to section \ref{sec:simulation_protocol_choice}, and helped edit the paper.
MRS: Helped create figure~\ref{fig:fig_what_is_lambda}, wrote sections describing choices for alchemical pathways and analysis for free energy calculations.
LNN: Helped write the simulation length, stopping conditions, and information saving section.
BA: Helped write the uncertainty estimation, stopping conditions, and output analysis sections and created figure ~\ref{fig:convergence_forward_reverse}.
JM: Contributed to sections~\ref{subsec:reproducible},~\ref{sec:step1},~\ref{} 4.2, 5.3, 6, 8.3.1, 8.3.2\
JDC: Wrote section~\ref{sec:decorrelating-samples} and~\ref{sec:automatic-equilibration-detection} discussed structure and design of the whole document, suggested Figs.~\ref{fig:fig_what_is_alchemy} and~\ref{fig:fig_sampling_scheme} \todo[inline, color={red!20}]{@JDC check info}
HX: Contributed section~\ref{subsec:accuracy}, to section~\ref{sec:relative-fe-protocol}, and to section~\ref{sec:are-they-good}.
